---
title: "Model calibration (model-assisted weighting)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model calibration (model-assisted weighting)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
has_ranger <- requireNamespace("ranger", quietly = TRUE)
```

Ordinary calibration forces the weighted sample to match known population totals
of *auxiliary* variables. Model calibration (Wu and Sitter, 2001) goes further:
it uses a working model for the *outcome* to build a better calibration variable.
`step_model_calibration()` implements it, and — less commonly — lets you keep the
ordinary consistency constraints on the auxiliaries in the **same** step.

## The idea

You do not know the population total of an outcome `y` — estimating it is the
goal. But you can fit a working model for `y` on auxiliaries `x` known for the
whole population, and calibrate on the model's **predictions**:

1. Fit `y_hat = m(x)` on the sample (weighted by the current weights).
2. Predict `y_hat` for every unit, in the **sample** and in the **population**
   (the predictors `x` are known throughout the frame).
3. Calibrate so the weighted sample total of the predictions equals their
   population total:

```
sum_{i in sample} w_i * y_hat_i  =  sum_{i in population} y_hat_i.
```

The target is the population total of the *fitted values*, which is computable —
not the unknown total of `y`. If the model predicts `y` well, the estimator of
the `y` total is efficient, because its error depends on the residuals
`y_i - y_hat_i` rather than on `y` itself.

## In weightflow, with a random forest

`step_model_calibration()` takes consistency auxiliaries (`x_formula`, with known
population totals), a named list of working `models` from `y_model()`, and the
`population` frame used to compute the prediction totals. Here the working model
is a **random forest**, which captures nonlinear structure in the predictors
without specifying it.

```{r recipe, eval = has_ranger}
spec <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = "region") |>
  step_model_calibration(
    x_formula  = ~ region + sex,                       # consistency constraints
    models     = list(income = y_model(income ~ age + sex + region,
                                       engine = "forest")),
    population = population)
fitted <- prep(spec)
```

The model is fitted on the **active** units only (the respondents, where `income`
is observed), weighted by the weights entering the step.

## Validation: the constraints are met exactly

Both kinds of constraint hold afterwards: the weighted totals of the consistency
auxiliaries match their known population totals, and the weighted total of the
income predictions matches its population total. The step's diagnostics show
target versus achieved:

```{r check, eval = has_ranger}
fitted$steps[[2]]$diagnostics
```

The `X (consistency)` rows are the region and sex totals; the `y (model)` row is
the population total of the forest predictions. Each `achieved` equals its
`target`.

## Does it help the estimate?

Because the weights are calibrated on the income predictions, the weighted
estimate of the income total tracks the true population total:

```{r estimate, eval = has_ranger}
est_total  <- sum(fitted$final_weight * sample_survey$income, na.rm = TRUE)
true_total <- sum(population$income)
c(estimated = est_total, population = true_total,
  rel_error = (est_total - true_total) / true_total)
```

The estimator stays design-based (`sum(w_i y_i)`); the model only shapes the
weights to make it efficient. A poor model does not bias the result — it just
yields a less efficient one.

## What you need: unit-level auxiliaries

This is the practical catch that sets model calibration apart. Raking,
post-stratification and GREG need only the population **totals** (or margins) of
the auxiliaries — numbers such as "1570 people in the North, 2311 women" — which
official statistics routinely publish. Model calibration needs more: the
auxiliaries **at the unit level for the whole population**, because step 2
predicts `y_hat` for every population unit before summing. You cannot get
`sum(y_hat_i)` from the totals of `x` alone (unless the model is linear, in which
case you are back to GREG).

That difference is visible in the arguments: `step_calibrate()` takes `margins`
or `totals` (vectors of totals), whereas `step_model_calibration()` takes
`population` — a data frame with **one row per population unit** carrying the
predictors. So model calibration applies when you have a register or frame with
microdata (a population register, administrative records, a census microdata
file, a large reference survey) — the "complete auxiliary information" of the
Wu-Sitter title means *unit-level for the whole population*, not just aggregates.
When you only have published totals, use raking, post-stratification or GREG.
The consistency constraints in the hybrid still calibrate to the `x` totals
(obtained by summing `population`); it is the model term that requires the
microdata.

## Combining the model with consistency constraints

The distinctive part of the step is the **hybrid**: it calibrates on the model
predictions *and* on the raw auxiliary totals (`x_formula`) at once. This pairing
matters precisely when the model is **nonlinear**. With a linear working model
the predictions lie in the span of `x`, so adding them is redundant and the
estimator reduces to ordinary linear calibration — the generalized regression
(GREG) estimator. With a forest (or any nonlinear model), the predictions carry
information beyond the linear `x` terms, so the two sets of constraints are
complementary: the `x` totals give **internal consistency / benchmarking** (the
weights reproduce the published region and sex totals), while the model term adds
**efficiency** from the nonlinear fit.

## Relationship to the literature

Model calibration was introduced by Wu and Sitter (2001), who showed it handles
any linear or nonlinear working model and reduces to the Deville-Särndal
calibration / GREG estimator in the linear case. Wu (2003) proved its optimality
within a class of calibration estimators, and Montanari and Ranalli (2005)
extended it to nonparametric models, close in spirit to the forest used here.
Särndal (2007) reviews the approach within the general calibration framework.

Combining the model term with the ordinary consistency constraints is supported
by the fact that calibration accommodates an arbitrary number of constraints
(Deville and Särndal, 1992); the practical motivation — efficiency from the model
*plus* benchmarking to published totals — is standard. What is uncommon is a
ready-to-use implementation: the major R survey packages calibrate on auxiliary
totals (GREG, raking, post-stratification) but do not fit an outcome model and
calibrate on its predictions, let alone with flexible engines combined with
consistency constraints. `step_model_calibration()` packages exactly that. It is
an accessible implementation of an established idea, not a new method.

**References.** Wu & Sitter (2001), *JASA* 96(453), 185–193. Wu (2003),
*Biometrika* 90(4), 937–951. Montanari & Ranalli (2005), *JASA* 100, 1429–1442.
Särndal (2007), *Survey Methodology* 33(2), 99–119. Deville & Särndal (1992),
*JASA* 87(418), 376–382.
