---
title: "Nonprobability samples: pseudo-weights by volunteering propensity"
output: rmarkdown::html_vignette
vignette: >
  %¥VignetteIndexEntry{Nonprobability samples: pseudo-weights by volunteering propensity}
  %¥VignetteEngine{knitr::rmarkdown}
  %¥VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
```

A nonprobability sample (a web panel, an opt-in survey, an opportunity source)
has no known inclusion probabilities, so its design weights do not exist and the
units enter through an unknown self-selection mechanism. Estimating population
quantities from it requires borrowing structure from a probability **reference**
sample. This article shows how `weightflow` builds inverse-propensity
pseudo-weights for a nonprobability sample, and is honest about what they can and
cannot fix.

## Notation and the two samples

Let $U = ¥{1, ¥dots, N¥}$ be the population, $Y = ¥sum_{i ¥in U} y_i$ the total of
the study variable and $¥bar Y$ its mean. We have two samples from $U$:

* $A$: a probability **reference** sample of size $n_A$, with weights $w_i^A$.
  We write $w_i^A$ (not a design weight specifically) because the reference
  weights may be design weights or already-calibrated weights; either way they
  make $A$ represent the population.
* $B$: the **nonprobability** sample of size $n_B$, with no known weights. Its
  units self-select (volunteer) into $B$.

Both samples observe a common set of covariates $¥mathbf{x}_i$ (and, in $B$, the
outcome $y_i$).

## Volunteering propensity and the key assumption

The true **volunteering propensity** of a unit is

$$¥phi_i^B = P(i ¥in B ¥mid ¥cdot),$$

the probability that unit $i$ self-selects into $B$. This is unknown and, in
general, may depend on anything, including the outcome $y_i$ itself: people
volunteer for reasons often related to what we are trying to measure. That makes
self-selection, in general, **non-ignorable** (unlike classical nonresponse,
where assuming ignorability given covariates is often defensible).

To make progress, the quasi-randomization approach **assumes** that volunteering
is ignorable given the covariates,

$$P(i ¥in B ¥mid ¥mathbf{x}_i, y_i) = P(i ¥in B ¥mid ¥mathbf{x}_i),$$

and, under this assumption, estimates the propensity as a function of the
covariates alone, $¥hat¥phi_i^B = ¥hat¥phi^B(¥mathbf{x}_i)$. The dependence on
$¥mathbf{x}$ is a property of the **estimate** (inherited from the assumption),
not a known property of the true $¥phi_i^B$. When the assumption fails (the true
volunteering depends on $y$ beyond $¥mathbf{x}$), the correction will be only
partial, as we will see.

## Estimating the propensity by pooling A and B

The propensity is estimated by stacking the two samples and modelling membership:
the units of $B$ get an indicator $1$, the units of $A$ an indicator $0$, and a
model predicts $¥hat¥phi^B(¥mathbf{x}_i)$ from the common covariates. The
reference weights $w_i^A$ enter this model so that the propensity targets the
population rather than the (unweighted) reference. The pseudo-weight of a
nonprobability unit is then the inverse-odds form (Elliott and Valliant, 2017),

$$w_i^B = w_i^{0}¥,¥frac{1 - ¥hat¥phi_i^B}{¥hat¥phi_i^B},$$

where $w_i^0$ is an initial weight, rescaled so the pseudo-weights represent the
population.

## A recipe for a nonprobability sample

The bundled `nps_sample` is a small volunteer sample (its participation depends
on age, sex, region **and** income), and `reference_sample` is a large
probability sample with design weights `pw_ref` summing to the population. We
start the recipe with `weighting_spec_nonprob()` (the sample has no design
weights), set the initial weights, and adjust by the volunteering propensity.

```{r recipe}
fitted <- weighting_spec_nonprob(nps_sample) |>
  step_initial_weights(method = "unity", pop_size = 1e6) |>
  step_pseudo_weight(reference = reference_sample,
                     reference_weights = "pw_ref",
                     formula = ‾ region + sex + age,
                     engine = "logit") |>
  prep()

fitted$steps[[2]]$diagnostics
```

### Initial weights

`step_initial_weights()` fixes the starting weights $w_i^0$ before the
adjustment. Several conventions from the literature are available: `"unity"`
(unit weights, optionally normalized to the population size, which performs well
in simulations), `"naive"` (unit weights, no normalization), `"constant"` (a
fixed value such as $N/n$), and `"ipa"` (inclusion-probability adjusted: each
nonprobability unit borrows the design weight of the most similar reference
unit).

### Flexible engines and cross-fitting

The propensity engine can be logistic regression or a machine-learning learner
(`"tree"`, `"forest"`, `"boost"`), useful when volunteering depends on the
covariates nonlinearly. As with the nonresponse step, flexible learners can
overfit the pooled membership, so `crossfit = K` estimates each propensity
out-of-fold. The `pooling` argument controls how the two samples' weights enter
the model (`"unweighted_nps"`, `"weighted"`, `"pseudo_lik"`).

## Does it reduce the bias?

The point of the pseudo-weights is to move the nonprobability estimate toward the
truth. Compare the naive (unweighted) mean of income in $B$, the pseudo-weighted
mean, and the true population mean:

```{r compare}
w_pseudo <- fitted$final_weight
naive    <- mean(nps_sample$income)
pseudo   <- sum(w_pseudo * nps_sample$income) / sum(w_pseudo)
truth    <- 22264   # known for this synthetic population

c(naive = round(naive), pseudo = round(pseudo), truth = truth)
```

The pseudo-weights pull the estimate down from the naive value toward the truth,
correcting a large part of the selection bias.

## The honest part: residual bias and non-ignorability

The correction is **partial**, and this is the central lesson, not a defect. In
this example volunteering was made to depend on `region`, `sex`, `age` **and**
directly on `income`. The propensity model only sees the covariates
`region + sex + age`; it cannot capture the part of the self-selection driven by
`income` itself, because the outcome does not enter the propensity (and is not
even observed in the reference). That residual is exactly the footprint of
**non-ignorable** volunteering: the assumption
$P(i ¥in B ¥mid ¥mathbf{x}, y) = P(i ¥in B ¥mid ¥mathbf{x})$ does not hold, so a
gap to the truth remains.

The practical implications are the ones official statistics care about. The
quality of the adjustment depends entirely on having covariates that explain the
self-selection; pseudo-weights are not a black box that removes bias, but a
device that removes the bias **captured by the covariates**. Choosing
$¥mathbf{x}$ that is related both to volunteering and to the outcomes is what
makes the method work, and no amount of modelling recovers a mechanism that the
covariates do not contain.

This residual is also what motivates **doubly robust** estimation, which pairs
the propensity model with a model for the outcome: if either model is correctly
specified, the estimator is approximately unbiased, giving a second chance to
remove the bias the propensity alone leaves behind.

## References

- Elliott, M. R., & Valliant, R. (2017). Inference for nonprobability samples.
  *Statistical Science*, 32(2), 249-264.
- Valliant, R., & Dever, J. A. (2011). Estimating propensity adjustments for
  volunteer web surveys. *Sociological Methods & Research*, 40(1), 105-137.
- Wang, L., Valliant, R., & Li, Y. (2021). Adjusted logistic propensity weighting
  methods for population inference using nonprobability volunteer-based cohorts.
  *Statistics in Medicine*, 40(24), 5237-5250.
- Chen, Y., Li, P., & Wu, C. (2020). Doubly robust inference with nonprobability
  survey samples. *JASA*, 115(532), 2011-2021.
- Beaumont, J.-F. (2020). Are probability surveys bound to disappear for the
  production of official statistics? *Survey Methodology*, 46(1), 1-28.
