---
title: "Nonresponse: weighting classes and propensities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonresponse: weighting classes and propensities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
has_rpart  <- requireNamespace("rpart",  quietly = TRUE)
has_ranger <- requireNamespace("ranger", quietly = TRUE)
```

Nonresponse adjustment inflates the weights of respondents so they also
represent the nonrespondents. `step_nonresponse()` offers two routes: weighting
classes and response propensity models. This vignette explains both, when each
is preferable, and how they are estimated.

Throughout, only **active** units (weight > 0) take part, so cases already
dropped earlier in the recipe (unknown eligibility, ineligible) are excluded
automatically.

## Weighting classes

Units are grouped into cells defined by `by`, and within each cell the
respondents absorb the weight of the nonrespondents. The adjustment factor in a
cell `c` is

```
f_c = ( sum of weights of all active units in c ) /
      ( sum of weights of the respondents in c )
```

Each respondent's weight is multiplied by `f_c`; nonrespondents go to zero.

```{r wc}
wf <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = "region") |>
  prep()
summary(wf)
```

**Validation.** By construction the total weight is preserved *within each cell*
(the nonrespondents' weight is moved to the respondents, not lost). So the
weighted total per region after the step equals the base-weight total before it:

```{r wc-check}
before <- tapply(sample_survey$pw,  sample_survey$region, sum)
after  <- tapply(wf$final_weight,   sample_survey$region, sum)
round(cbind(before, after, diff = after - before), 6)
```

The differences are zero: weighting classes redistribute, they do not create or
destroy weight.

## Response propensities

Instead of cells, the probability of responding is modelled from auxiliaries
known for respondents and nonrespondents alike:

```
p_i = P(respond | x_i)
```

The model is fitted **on the active units, weighted by the current weights**, and
two routes follow. With `num_classes = NULL`, each respondent is weighted by the
inverse propensity, `w_i* = w_i / p_i`. With an integer `num_classes`, units are
grouped into that many propensity classes and a weighting-class adjustment is
applied within each, which is more robust to a misspecified model.

### Logistic regression

```{r prop-logit, warning = FALSE}
wf <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "propensity",
                   formula = ~ region + sex + age, engine = "logit",
                   num_classes = 5) |>
  prep()
summary(wf)
```

Because the model is fitted with survey weights, a logistic fit may print a
"non-integer #successes" message: that is expected for a weighted binomial fit
and does not affect the estimated propensities.

### Trees and forests

The same propensity can be estimated with a regression tree (`engine = "tree"`,
package **rpart**) or a random forest (`engine = "forest"`, package **ranger**),
which capture nonlinearities and interactions without specifying them. More
flexibility is not free, though: a very flexible model can overfit the response
and produce more dispersed adjustment factors, which *raises* the variance of
the weights (a higher design effect). Compare the `deff` after each engine
below — the forest typically yields the largest, the weighting classes the
smallest.

```{r prop-tree, eval = has_rpart}
wf <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "propensity",
                   formula = ~ region + sex + age, engine = "tree",
                   num_classes = 5) |>
  prep()
design_effect(wf$final_weight)$deff
```

```{r prop-forest, eval = has_ranger}
wf <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "propensity",
                   formula = ~ region + sex + age, engine = "forest",
                   num_classes = 5) |>
  prep()
design_effect(wf$final_weight)$deff
```

## Person or household level

Nonresponse can occur at the person level (within a reached household) or at the
household level (the whole household is not reached). The `cluster` argument
moves the adjustment to the household: each household counts once with its
weight, and the redistribution (or the propensity model) is done over
households, then assigned to their members.

```{r cluster}
wf <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = "region", cluster = "household_id") |>
  prep()
design_effect(wf$final_weight)$deff
```

The level is dictated by what is known about the nonrespondents: household
auxiliaries and a whole-household outcome call for `cluster`; person-level
auxiliaries within reached households do not. Note that the effective sample
size drops more at the household level, since households — not persons — are the
independent units being adjusted.

## Which to use

Weighting classes need categorical auxiliaries and enough respondents per cell;
they are simple and transparent. Propensity models handle continuous predictors
and many auxiliaries at once, and the tree/forest engines relax functional-form
assumptions. Using propensity **classes** (`num_classes`) rather than the direct
`1 / p` keeps the adjustment stable when the model is imperfect, at the cost of
some efficiency. In all cases, model the response on auxiliaries that are both
predictive of responding and related to the survey outcomes.
