---
title: "Documenting and auditing the weights: the quality report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting and auditing the weights: the quality report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
```

`report_weighting()` turns a prepped recipe into a single, self-contained HTML
report that documents how the weights were built and how they behave. It follows
the logic of an official-statistics quality report (GSBPM sub-process 5.6 and the
ESS SIMS/ESMS metadata concepts) and needs no external dependency: every plot is
inline SVG, so the file opens anywhere and can be archived or emailed as is.

## A recipe and the report

We build a short cascade on the bundled `sample_survey` and generate the report.
Passing `open = FALSE` writes the file without launching a browser, which is what
you want inside a script; the function returns the path invisibly.

```{r report}
fit <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = "region") |>
  step_calibrate(method = "raking",
                 margins = list(region = c(table(population$region)),
                                sex    = c(table(population$sex)))) |>
  prep()

path <- report_weighting(fit, open = FALSE)
```

Open `path` in a browser to see the report. The rest of this vignette describes
what each part of it says.

## What the report contains

The report is a sequence of cards, in reading order.

Executive summary. A prose summary of the cascade with the headline design
effect, effective sample size and R-indicator, a plain-language interpretation of
the final design effect, and a status checklist (convergence, extreme weights,
replicate weights, alerts). Anything that needs attention (for example a step
that did not converge, or a quality alert) is collected here.

Fieldwork outcomes (AAPOR). When the recipe has eligibility or nonresponse
steps, a card reconstructs the disposition of every case (ineligible, unknown
eligibility, eligible respondent, eligible nonrespondent) and reports the
eligibility rate together with the response rate in three AAPOR variants, from
the most to the least conservative treatment of the unknown-eligibility cases
(RR1, RR3/CASRO, RR5), both unweighted and base-weighted.

Per-stage summary. A table with the active units, sum of weights, CV, Kish
design effect and effective sample size at each stage; an inline chart of the
design effect across stages; and a per-step impact table (the change in deff and
CV at each step and its share of the total change).

Weight distribution and per-step detail. The distribution of the final weights,
and a collapsible section with one card per step: the requested parameters, the
diagnostics (for calibration, the target and achieved totals with their relative
difference), and a before/after scatter of the weights.

## Reference metadata

The `metadata` argument adds a header card aligned to the SIMS/ESMS concepts.
Recognised keys get a proper label; any other key is shown as given.

```{r metadata}
path <- report_weighting(
  fit, open = FALSE,
  metadata = list(
    survey           = "Living Conditions Survey (illustrative)",
    reference_period = "2024",
    geography        = "National",
    producer         = "National Statistical Office",
    frame            = "Master sample of dwellings",
    totals_source    = "Population projections (2023)",
    totals_date      = "2024-06-30"))
```

## Language and the variance card

The narrative is available in English (`lang = "en"`, the default) or Spanish
(`lang = "es"`). If you also computed replicate weights, pass them through
`replicates` to add a card documenting the variance replication design: method,
number of replicates, strata and PSUs per stratum, lonely-PSU handling, cores and
run time, with a note when few PSUs per stratum favour the jackknife.

```{r variance-card, eval = FALSE}
spec <- weighting_spec(sample_survey, base_weights = pw) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = "region")
jk <- jackknife_weights(spec, strata = "region", psu = "psu")
report_weighting(prep(spec), replicates = jk, lang = "es", open = FALSE)
```

The variance card documents the design of the replication. The standard errors,
CVs and confidence intervals of specific estimates are computed downstream with
the `survey` or `srvyr` package (see the Variance estimation article).
