---
title: "Rotating panels: net change, chaining and gross flows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rotating panels: net change, chaining and gross flows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
set.seed(20260910)
```

A continuous household survey with a rotating design measures the same units more
than once. That is what makes the **change** between two periods far more precise
than the levels themselves -- the shared sample cancels part of the sampling error --
and it is also what makes the change **harder** to estimate: the two samples are not
independent, so `V(change)` is not `V1 + V2`.

This vignette covers the three things a rotating panel makes possible, in the order
an office needs them:

1. the **net change** between two periods, with its honest variance;
2. **chaining** period after period the way production actually runs, one month at a
   time, with only a small object traveling between runs;
3. the **gross flows** -- who moved between states -- which need a longitudinal weight.

The package's panel layer follows the methodology of Statistics Canada's Labour Force
Survey (cat. 71-526-X, sec. 7.2.2) for the coordinated replication, and ECLAC's
household-survey manual (chapters XVI-XVII) for the longitudinal weight and the flows.

## The structure comes first

Before any weighting, `panel_design()` reads the unit x wave crossing and describes
what is actually there. It computes nothing about weights.

```{r design}
pd <- panel_design(panel_ine, unit = c("id_hogar", "nper"), wave = "ola",
                   rotation_group = "grupo_rotacion", pattern = "6")
pd
```

Two things to read in that output.

The **overlap matrix** is observed, from the data. The **profile implied by the
pattern** is theoretical, derived from the rotation calendar. The declared `pattern`
is therefore *verification*, not configuration: when the observed overlap falls below
what the design implies, the linkage key is suspect, and `PN-01` says so. Ordinary
attrition pulls the observed overlap down a little; a broken key pulls it down a lot.

The pattern is parsed into the whole profile, not just the adjacent lag, because the
informative lag is not always lag 1:

```{r profiles}
prof <- function(p) round(weightflow:::.wf_pattern_overlap(p)$profile[1:5], 3)
rbind(`6        (Canada LFS, ECH Uruguay)` = prof("6"),
      `4(0)1    (ECLAC ch. XVI example)`   = prof("4(0)1"),
      `2-(2)-2  (Chile ENE, Italy)`        = prof("2-(2)-2"),
      `4-8-4    (US CPS)`                  = prof("4-8-4"),
      `1(2)5    (PNAD Continua)`           = prof("1(2)5"))
```

`2-(2)-2` shares **no** sample at lag 2 and half of it at lag 4 -- as much as at lag 1 --
so a year-on-year change there needs as much coordination as a quarter-on-quarter one.
The PNAD Continua design `1(2)5` shares nothing with the adjacent quarter at all: a design built around
"the previous period" would be exactly backwards for it. Both notations are accepted
and describe the same calendar: `"2-(2)-2"` and `"2(2)2"` are the same design.

## Net change between two periods

The coordinated bootstrap resamples **PSUs**, and a PSU present in both waves is
resampled the same way in both. That is what lets the overlap show up as covariance.

```{r change}
w1 <- subset(panel_ine, ola == 1 & disp == "R")
w2 <- subset(panel_ine, ola == 2 & disp == "R")
rec <- function(d) weighting_spec(d, base_weights = w_base) |>
  step_nonresponse(respondent = disp == "R", by = "sexo")

wb <- wave_bootstrap(list(T1 = rec(w1), T2 = rec(w2)), replicates = 100,
                     strata = "estrato", psu = "psu", seed = 1, progress = FALSE)
change_mean(wb, "desocupado")
```

`rho` is the correlation the overlap induces and `deff_change` is `V / (V1 + V2)`:
the ratio between the variance reported here and what an office would report if it
treated the two periods as independent samples. Ignoring the overlap does not give a
conservative answer -- it gives a wrong one, in either direction depending on the sign
of the covariance.

For a combination over more than two waves -- a rolling quarter, an annual average --
`panel_estimate()` takes an arbitrary contrast:

```{r contrast, eval = FALSE}
panel_estimate(wb, mean_of("desocupado"), contrast = c(-1, 1))   # the net change
panel_estimate(wb, mean_of("desocupado"))                        # the average
```

## Chaining: how production actually runs

`wave_bootstrap()` wants every wave at once. An office does not have them: it
publishes month `t` weeks after month `t-1`, and the twelfth month of the year cannot
wait for the first eleven to be reprocessed. `wave_step()` runs **one period at a
time** and hands the next period a small object -- the *carry* -- that is all it needs.

```{r chain, eval = FALSE}
# period 1: nothing to coordinate with yet
s1 <- wave_step(rec(w1), estimands = EST, replicates = 500,
                strata = "estrato", psu = "psu", period = "2026-01", seed = 1)
saveRDS(wave_carry(s1), "carry/2026-01.rds")

# period 2, weeks later, in a fresh session
prev <- readRDS("carry/2026-01.rds")
s2 <- wave_step(rec(w2), previous = prev, estimands = EST, replicates = 500,
                strata = "estrato", psu = "psu", period = "2026-02", seed = 2)
s2$weights   # the cross-sectional weights the office publishes
s2$change    # the net change against 2026-01, with rho and deff_change
s2$strata    # the coordination diagnostic, stratum by stratum
saveRDS(wave_carry(s2), "carry/2026-02.rds")
```

Three properties worth stating plainly.

**The cross-sectional weights are untouched.** `s2$weights` is identical to
`prep(spec)$final_weight`. Coordination adds the change and the carry; it never moves
the point estimate the office publishes.

**`previous` is a list, not a file.** Which earlier periods share sample with this one
is decided by the rotation calendar, not by proximity: with `2-(2)-2` the useful lags
are 1, 3, 4 and 5, and lag 2 is empty. Each PSU inherits its multiplicity from the
most recent carry that contains it, so gaps and returning cohorts resolve themselves.

**The coordination is exact when the stratum keeps its size.** Coordination transfers
the PSU *multiplicities*, and when `n_h` is unchanged the transfer is a permutation --
case (ii) of the LFS methodology -- so no replicate needs adjusting. `$strata` reports
the case and the share of replicates that closed without adjustment, per stratum. That
column is the quality indicator that decides whether a figure is publishable.

### Combinations over the chain

`wave_step()` reports pairwise changes. A rolling quarter is not pairwise, and
`wave_contrast()` estimates any linear combination straight from the saved carries:

```{r wavecontrast, eval = FALSE}
tr <- lapply(c("2026-01", "2026-02", "2026-03"),
             \(m) readRDS(sprintf("carry/%s.rds", m)))

wave_contrast(tr, "unemployment_rate")                          # rolling quarter
wave_contrast(tr, "unemployment_rate", contrast = c(-1, 0, 1))  # T3 - T1
```

This works without the waves being in memory because each carry stores the `R`
replicate values of every declared estimand, and those replicates are **paired across
periods** -- the coordination transferred the multiplicities. Stacking them recovers
the full covariance matrix, and any contrast follows from it.

### The composite estimator

When the recipe ends in `step_cre()`, the level of period `t` depends on control
totals **estimated** with the previous wave. Treating them as known constants makes
the variance anticonservative, so replicate *b* of period `t` rebuilds `Zhat*` from
replicate *b* of period `t-1`. That is why a chain with `step_cre()` needs the "fat"
carry, which weighs a few hundred KB instead of a few dozen: it must carry the
previous period's replicate weights. `$n_cre_injected` and `$n_cre_skipped` are the
audit that it actually happened.

## Gross flows: who moved

A net change is a difference of aggregates. It cannot distinguish an immobile
population from one where equal numbers enter and leave employment. For that you need
the **longitudinal** file and a longitudinal weight.

```{r flows}
wide <- panel_merge(list(T1 = subset(panel_ine, ola == 1),
                         T2 = subset(panel_ine, ola == 2)),
                    by = c("id_hogar", "nper"), require = "all")

lw <- weighting_spec(wide, base_weights = w_base_T1) |>
  step_drop_ineligible(disp_T2 == "OS", reason = "left the target population") |>
  step_attrition(respondent = disp_T2 == "R", method = "propensity",
                 formula = ~ edad_T1 + sexo_T1) |>
  prep()

transition_matrix(lw, from = "condicion_T1", to = "condicion_T2", format = "row")
```

`boot_transition()` adds a standard error per cell, and `boot_flows()` gives the same
information as population totals plus the net flows `i->j` minus `j->i` and the
margins -- how many started in each state, ended in each, stayed, and moved.

A caveat about the illustration, not the method: the four shipped panel datasets are
synthetic, and in all of them **inactivity is an absorbing state** -- nobody enters or
leaves it. So the matrix above shows movement only between employment and unemployment,
and the `inact` row and column are degenerate. The mechanics are the point here; on real
microdata the transitions in and out of inactivity are usually the interesting ones.

Note the order of the two steps, which is not cosmetic. **Leaving the target
population is not nonresponse.** Someone who died or emigrated is removed from the
universe with `step_drop_ineligible()`; their weight is not redistributed to anyone.
Someone who is still in the universe but did not answer is attrition, and their weight
*is* redistributed, among the units that remain. Collapsing the two inflates the
population.

### Three ECLAC conventions the package cannot enforce

These are decisions the analyst makes; the package has no way to check them, so they
are stated here (ECLAC, ch. XVI).

**The longitudinal population** is the units that were in the target population at the
first period **and stayed** through the last. Those who left are out; those who
*entered* later are out too. A panel adds no elements over time.

**Calibrate to the first period's totals.** The manual is explicit that the auxiliary
totals must represent the population of the *first* period, because a panel that adds
no elements is representative only of the period in which it was selected. Passing the
later period's projections is a silent error: the recipe converges, the totals close,
and the weights represent a population the sample never had a chance to cover. A
vector of control totals carries no label saying which period it is, so nothing but
you can catch this.

**Cross-sectional estimates from a longitudinal file are reference only.** They will
not match the published cross-sectional figures, and they should not: the target
population of the panel combination is not the target population of the cross-section.
Use the longitudinal weight for flows, transitions and durations; use the
cross-sectional weight for levels.

## Where to look next

This vignette is the entry point; three companions go deeper into the pieces it
uses.

* `vignette("coordinated-replication")` -- what actually travels between waves,
  the four coordination cases, and how to read the `$strata` diagnostic that
  decides whether a change is publishable.
* `vignette("composite-estimation")` -- `step_cre()` in full: what the composite
  estimator buys, and why its control totals make its variance a special case.
* `vignette("panel-longitudinal")` -- the **pure** panel: attrition over many
  waves, the longitudinal weight, and gross flows.

For the reference pages: `?wave_step` and `?wave_carry` for the chaining engine,
`?wave_contrast` for combinations over a chain, `?panel_design` for the structure
layer, `?transition_matrix` for the flows, and `?report_panel` for a quality report
of a panel run. `vignette("variance-estimation")` covers the single-period
bootstrap this one builds on, and `vignette("validation")` checks the change
variance against an analytic estimator from a different family.
