---
title: "Validation against the survey package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation against the survey package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
has_survey <- requireNamespace("survey", quietly = TRUE)
```

weightflow's calibration is meant to reproduce the established results of the
`survey` package on the methods they share (raking, post-stratification and
linear GREG calibration) while adding the staged cascade (eligibility,
nonresponse, selection) and a recipe-aware bootstrap on top. This vignette
checks that agreement directly: on the same starting weights and the same
control totals, the two packages return the same weights.

To make every unit comparable one-to-one, the recipes below use **only** the
calibration step (no dropping or nonresponse), so no rows are removed.

```{r data}
d <- sample_survey
N <- nrow(population)
```

## Post-stratification

Post-stratifying to the population counts of `region`: each region's weights are
rescaled so the weighted count matches the known total.

```{r poststratify, eval = has_survey}
library(survey)

# weightflow
wf <- weighting_spec(d, base_weights = pw) |>
  step_calibrate(method = "poststratify",
                 margins = list(region = c(table(population$region)))) |>
  prep()
w_wf <- wf$final_weight

# survey
des    <- svydesign(ids = ~1, weights = ~pw, data = d)
pr     <- data.frame(region = names(table(population$region)),
                     Freq = as.numeric(table(population$region)))
des_ps <- postStratify(des, ~region, pr)
w_sv   <- weights(des_ps)

c(max_abs_weight_diff = max(abs(w_wf - w_sv)))
```

## Raking

Raking (iterative proportional fitting) to the `region` and `sex` margins. We
tighten `survey`'s convergence so both solve the system to the same precision.

```{r raking, eval = has_survey}
# weightflow
wf <- weighting_spec(d, base_weights = pw) |>
  step_calibrate(method = "raking",
                 margins = list(region = c(table(population$region)),
                                sex    = c(table(population$sex)))) |>
  prep()
w_wf <- wf$final_weight

# survey (tight epsilon so it fully converges, like weightflow)
des   <- svydesign(ids = ~1, weights = ~pw, data = d)
ps    <- data.frame(sex = names(table(population$sex)),
                    Freq = as.numeric(table(population$sex)))
des_rk <- rake(des, list(~region, ~sex), list(pr, ps),
               control = list(epsilon = 1e-10, maxit = 100))
w_sv  <- weights(des_rk)

c(max_abs_weight_diff = max(abs(w_wf - w_sv)))
```

## Linear (GREG) calibration

Linear calibration to the totals of the design matrix of `~ region + sex`,
including the intercept (the population size `N`).

```{r greg, eval = has_survey}
totals <- colSums(model.matrix(~ region + sex, population))

# weightflow
wf <- weighting_spec(d, base_weights = pw) |>
  step_calibrate(method = "linear", formula = ~ region + sex, totals = totals) |>
  prep()
w_wf <- wf$final_weight

# survey
des     <- svydesign(ids = ~1, weights = ~pw, data = d)
des_cal <- calibrate(des, ~ region + sex, population = totals, calfun = "linear")
w_sv    <- weights(des_cal)

c(max_abs_weight_diff = max(abs(w_wf - w_sv)))
```

## Same estimates

The agreement carries over to estimates. A calibrated total of a survey outcome
matches between the two packages:

```{r estimate, eval = has_survey}
wf <- weighting_spec(d, base_weights = pw) |>
  step_calibrate(method = "raking",
                 margins = list(region = c(table(population$region)),
                                sex    = c(table(population$sex)))) |>
  prep()
total_wf <- sum(wf$final_weight * d$employed, na.rm = TRUE)

des    <- svydesign(ids = ~1, weights = ~pw, data = d)
des_rk <- rake(des, list(~region, ~sex), list(pr, ps),
               control = list(epsilon = 1e-10, maxit = 100))
total_sv <- as.numeric(svytotal(~employed, des_rk, na.rm = TRUE))

c(weightflow = total_wf, survey = total_sv, difference = total_wf - total_sv)
```

## What weightflow adds

The point of agreement is trust: where the methods overlap, weightflow returns
exactly what `survey` does. On top of that shared core, weightflow contributes
the **staged cascade** (unknown eligibility, ineligible dropping,
within-household selection, and person- or household-level nonresponse, each as
a pipeable step with diagnostics) and a **bootstrap that re-applies the whole
recipe** on each replicate, so the variance reflects every adjustment (see the
*Variance estimation* article). For design-based inference you can always export
the final weights back to `survey`/`srvyr`.
