---
title: "Variance estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
has_survey <- requireNamespace("survey", quietly = TRUE)
has_srvyr  <- requireNamespace("srvyr", quietly = TRUE) &&
              requireNamespace("dplyr", quietly = TRUE)
```

weightflow computes weights and also estimates their variances. This vignette
shows two ways to obtain standard errors from a weightflow recipe, and how they
relate.

Throughout, $U$ is the population and $s$ the sample; $w_i$ is the final weight of
unit $i$; and a population total is written $Y = \sum_{i \in U} y_i$, estimated by
$\hat Y = \sum_{i \in s} w_i\,y_i$. The sample is drawn in clusters: primary
sampling units (PSUs) nested in strata.

## Why the adjustments matter for variance

A weighting recipe rarely stops at the design weight. It redistributes unknown
eligibility, drops out-of-scope units, adjusts for nonresponse and calibrates to
known totals. Each of those stages is *estimated from the sample*, so each one
adds (or, for calibration, often removes) variability.

A linearization that takes the final weights as fixed and applies the
ultimate-cluster formula ignores that the nonresponse and calibration steps were
themselves estimated. The cleanest way to account for them is to **re-run the
whole recipe on each replicate**, so the replicate weights carry the variability
of every stage.

## Method 1: a PSU bootstrap that re-applies the recipe

`bootstrap_weights()` resamples primary sampling units (PSUs) with replacement
within strata and re-runs the recipe on each replicate. Pass the **inert**
recipe (do not call `prep()` first): the bootstrap preps it once per replicate.

```{r recipe}
dat <- sample_one
dat$age_grp <- cut(dat$age, c(0, 30, 45, 60, Inf),
                   labels = c("18-30", "31-45", "46-60", "60+"))

spec <- weighting_spec(dat, base_weights = pw) |>
  step_unknown_eligibility(unknown = unknown_elig, by = "region",
                           cluster = "household_id") |>
  step_drop_ineligible(ineligible = ineligible) |>
  step_nonresponse(respondent = hh_responded, method = "weighting_class",
                   by = "region", cluster = "household_id") |>
  step_select_within(prob = p_within) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = c("region", "sex", "age_grp")) |>
  step_calibrate(method = "raking",
                 margins = list(region = c(table(population$region)),
                                sex    = c(table(population$sex))))

boot <- bootstrap_weights(spec, replicates = 200, strata = "region",
                          psu = "psu", seed = 2024, progress = FALSE)
boot
```

The multiplier is the **Rao-Wu rescaling bootstrap**. Consider a stratum $h$ with
$n_h$ PSUs, from which $m_h$ are drawn with replacement (by default $m_h = n_h -
1$). Let $t_{hi}^{*}$ be the number of times PSU $i$ is selected in a replicate.
Every unit in that PSU has its weight rescaled by

$$\lambda_{hi} = 1 - \sqrt{\tfrac{m_h}{n_h - 1}}
  + \sqrt{\tfrac{m_h}{n_h - 1}}\;\frac{n_h}{m_h}\,t_{hi}^{*},$$

so the replicate weight is $w_i^{*} = \lambda_{hi}\,w_i$. The factor has
expectation one over the resampling, $\mathbb{E}(\lambda_{hi}) = 1$, which keeps
each replicate design-unbiased, and the construction never turns it negative, so
the recipe can be re-prepped on every replicate without invalid weights. Whole
PSUs are kept together (every unit in a drawn PSU is retained), as the design's
clustering requires.

### Estimates with bootstrap standard errors

Writing $\hat\theta$ for the point estimate and $\hat\theta_b$ for its value on
replicate $b$ (each computed from the re-prepped replicate weights), the bootstrap
variance is the average squared deviation across the $B$ replicates,

$$\widehat{\operatorname{Var}}(\hat\theta)
  = \frac{1}{B} \sum_{b=1}^{B} \big(\hat\theta_b - \hat\theta\big)^2 .$$

```{r estimates}
boot_mean(boot,  "income")     # mean income
boot_total(boot, "employed")   # total employed
boot_mean(boot,  "employed")   # employment rate
```

For any other statistic, pass a function of the weights and the data to
`bootstrap_estimate()`:

```{r custom}
bootstrap_estimate(boot, function(w, d) {
  ok <- !is.na(d$income) & w > 0
  stats::median(rep(d$income[ok], times = round(w[ok])))   # weighted median (approx.)
})
```

## Method 2: hand the weights to the survey package

`as_svydesign()` builds an ultimate-cluster linearization design from a prepped
recipe. It is fast, but treats the calibration as fixed.

```{r survey, eval = has_survey}
fitted <- prep(spec)
des <- as_svydesign(fitted, ids = "psu", strata = "region")
survey::svymean(~income, des, na.rm = TRUE)
```

To keep the recipe's adjustments in the variance while still using survey, feed
it the bootstrap replicate weights from method 1:

```{r svrep, eval = has_survey}
rep_des <- as_svrepdesign(boot)
survey::svymean(~income, rep_des, na.rm = TRUE)
```

This matches `boot_mean(boot, "income")` exactly, because `as_svrepdesign()` sets
`scale = 1 / B`, `rscales = 1` and `mse = TRUE`.

## Replicate weights for a tidyverse workflow

`collect_replicate_weights()` attaches the point weight (`.weight`) and the
replicate weights (`rep_1` ... `rep_B`) to the active respondents, ready for
srvyr.

```{r srvyr, eval = has_srvyr}
df <- collect_replicate_weights(boot)
d_rep <- srvyr::as_survey_rep(df, weights = .weight,
                              repweights = dplyr::starts_with("rep_"),
                              type = "bootstrap", combined.weights = TRUE,
                              scale = 1 / attr(df, "R"), rscales = 1, mse = TRUE)
srvyr::summarise(d_rep, mean_income = srvyr::survey_mean(income, na.rm = TRUE))
```

## Which one to use

Use the **recipe-aware bootstrap** (method 1, in any of its three forms) when
the nonresponse and calibration steps are a meaningful part of the design and
you want their uncertainty reflected; it is the more honest variance. Use the
**linearization** (method 2) for a quick, well-understood standard error when
the adjustments are minor or you only need the design-and-clustering part.

A few practical notes. More replicates give a more stable bootstrap SE; 200 is
fine for exploration, 500-1000 for final figures. Each stratum needs at least
two PSUs to be resampled (single-PSU strata are left untouched, with a warning).
If a replicate leaves a calibration or weighting-class cell empty it is dropped
with a warning; coarser `by` cells make the bootstrap more robust.
