---
title: "Staged survey weighting: the adjustment logic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Staged survey weighting: the adjustment logic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(weightflow)
```

Survey weighting turns a sample into estimates for a population. It starts from
the design and then corrects, one stage at a time, for everything that pulls the
realized sample away from the target population: addresses that could not be
resolved, units out of scope, subsampling inside households, nonresponse, and
coverage. weightflow expresses that chain as a pipeline of explicit steps. This
vignette walks the logic of each stage, why it exists, and why the order
matters.

## The starting point: design weights

Under a probability design, unit `i` enters the sample with a known inclusion
probability `pi_i`. The design (base) weight is its inverse,

```
d_i = 1 / pi_i,
```

so the Horvitz-Thompson estimator of a total, `sum(d_i * y_i)` over the sample,
is unbiased for the population total. Every later stage multiplies this weight by
a correction factor; the recipe is just the product of those factors, applied in
order. We use the bundled multistage sample:

```{r data}
dat <- sample_one
dat$age_grp <- cut(dat$age, c(0, 30, 45, 60, Inf),
                   labels = c("18-30", "31-45", "46-60", "60+"))
```

## The cascade

### Unknown eligibility

Some sampled addresses are never resolved — you cannot tell whether they were
in scope. Discarding them would lose their share of the population; keeping them
as if eligible would overstate it. The standard fix redistributes their weight
over the resolved cases (eligible **and** ineligible) within cells, so the
resolved units carry the unresolved share. When the unknowns arrive without a
roster, this is done at the household level (`cluster`).

### Ineligible units

Resolved out-of-scope units (vacant addresses, non-residential) are simply
removed — their weight goes to zero with no redistribution. They must be present
during the unknown-eligibility step (so they absorb their share) and only then
dropped.

### Within-household selection

When one person is selected per household, that person represents all the
eligible persons there. The weight is multiplied by the inverse of the
within-household selection probability (`prob`), or by the number of eligibles
(`n_eligible`) under equal selection.

### Nonresponse

Not everyone responds. Under the assumption that response is ignorable given the
auxiliaries used (missing at random within cells or given the model), the
respondents are inflated to represent the nonrespondents — either with weighting
classes or a response-propensity model, at the person or household level. The
*Nonresponse* article covers this stage in detail.

### Calibration

Finally, auxiliary totals known for the whole population (here, region and sex
counts) are used to force the weighted sample to match them. Calibration reduces
coverage bias and, when the auxiliaries are related to the outcomes, lowers
variance; it is the weighting counterpart of the generalized regression (GREG)
estimator. The *Validation* article shows weightflow's calibration reproduces
the `survey` package.

```{r recipe}
fitted <- weighting_spec(dat, base_weights = pw) |>
  step_unknown_eligibility(unknown = unknown_elig, by = "region") |>
  step_drop_ineligible(ineligible = ineligible) |>
  step_nonresponse(respondent = hh_responded, method = "weighting_class",
                   by = "region") |>
  step_select_within(prob = p_within) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = c("region", "sex", "age_grp")) |>
  step_calibrate(method = "raking",
                 margins = list(region = c(table(population$region)),
                                sex    = c(table(population$sex)))) |>
  prep()
```

### Trimming, rounding, rescaling

Adjustments can create a few very large weights, which inflate the variance.
Trimming caps them (and redistributes the excess) to trade a little bias for a
worthwhile variance reduction; the bias-variance trade-off of trimming is a
classic topic (Potter). Rounding and rescaling are presentational: integer
weights, or weights that sum to the sample size (mean one) or to a chosen total.

A caveat on order: trimming after calibration preserves the overall total but
can perturb the calibrated margins slightly, so a rigorous workflow re-calibrates
after trimming.

```{r trim}
trimmed <- weighting_spec(dat, base_weights = pw) |>
  step_unknown_eligibility(unknown = unknown_elig, by = "region") |>
  step_drop_ineligible(ineligible = ineligible) |>
  step_nonresponse(respondent = hh_responded, method = "weighting_class",
                   by = "region") |>
  step_select_within(prob = p_within) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = c("region", "sex", "age_grp")) |>
  step_calibrate(method = "raking",
                 margins = list(region = c(table(population$region)),
                                sex    = c(table(population$sex)))) |>
  step_trim_weights() |>
  prep()
```

## Reading the cascade: the design effect

The Kish design effect, `deff = 1 + CV^2` (with `CV` the coefficient of
variation of the weights), measures how much unequal weighting inflates the
variance; the effective sample size is `n_eff = n_active / deff`. Each
adjustment tends to *raise* the design effect (it makes the weights more
unequal) and trimming brings it back down. The per-stage summary shows the whole
trajectory:

```{r summary}
summary(fitted)
```

And the effect of trimming on the final design effect:

```{r deff-compare}
c(no_trim = design_effect(fitted$final_weight)$deff,
  trimmed = design_effect(trimmed$final_weight)$deff)
```

## Why the order matters

The sequence is not arbitrary. Unknown eligibility comes first, so ineligibles
are still present to absorb their share before being dropped. Household
nonresponse precedes within-household selection (the household must be reached
before a person is selected), and person nonresponse follows it. Calibration
comes last, so it operates on weights that already reflect eligibility,
selection and response. Trimming and presentational steps close the pipeline.
weightflow makes this order explicit — it is simply the order in which you pipe
the steps — and records what each one did, which you can inspect with
`summary()`, `plot()` and `report_weighting()`. For standard errors that respect
the whole cascade, see the *Variance estimation* article.
