---
title: "A full weighting pipeline on a real household survey (ECH 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A full weighting pipeline on a real household survey (ECH 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7, fig.height = 4.4, dpi = 110)
```

This article puts the whole `weightflow` pipeline to work on a real survey: the
Uruguayan continuous household survey (Encuesta Continua de Hogares, ECH 2019),
released as open microdata by the national statistical office (INE). It doubles
as a stress test, since the working sample has tens of thousands of records and
the expanded population has more than three million.

The published microdata contain only the eligible respondents. A real survey
operation also faces ineligible units, units of unknown eligibility, and
nonresponse, and none of those carry survey answers. To exercise the full
pipeline we therefore induce these outcomes on the public data in a reproducible
(seeded) way, weight the survivors back to the population, and check how close we
land to a known truth.

The target parameter is the **poverty rate**
$$p = \frac{1}{N}\sum_{i \in U} y_i,$$
where $y_i = 1$ if person $i$ is poor and $U$ is the population. The estimator is
the weighted mean $\hat p = \sum_s w_i\,y_i / \sum_s w_i$, and the weights travel
through the cascade $w_i^0 \to w_i^{el} \to w_i^{nr} \to w_i^{cal} \to w_i$:
design base weight, unknown-eligibility adjustment, nonresponse adjustment,
calibration, and final trimming and rounding.

```{r libs}
library(weightflow)
library(dplyr)
library(tidyr)
library(haven)
library(archive)
library(ggplot2)

wf_violet <- "#3d3580"; wf_lav <- "#7a6ad0"; wf_amber <- "#e8941f"
```

## 1. Download the ECH microdata

The INE publishes, on its ANDA portal, the ECH 2019 person/household file and a
companion file with the stratum and primary sampling unit (PSU) of each
household. We download both and unpack them.

```{r download, eval = TRUE}
dir_ine <- "ech_data"
if (!dir.exists(dir_ine)) dir.create(dir_ine)

f1 <- tempfile(fileext = ".rar")
download.file("https://www4.ine.gub.uy/Anda5/index.php/catalog/715/download/956",
              f1, mode = "wb"); archive_extract(f1, dir = dir_ine)

f2 <- tempfile(fileext = ".rar")
download.file("https://www4.ine.gub.uy/Anda5/index.php/catalog/715/download/1154",
              f2, mode = "wb"); archive_extract(f2, dir = dir_ine)
```

## 2. Build the sample and the population

We keep the design identifiers (stratum, PSU, household), the calibration
auxiliaries (region, sex, age), the outcome (poverty), and the ECH person weight.
Each row is a person, and the design is two-stage: PSUs within strata, with the
household as the last stage.

A key modelling choice: we set the base weight equal to the **individual ECH
weight** `w_ech`. With this choice the base-weighted poverty rate reproduces the
population rate exactly, so the sample starts at the truth and any later gap is
caused by the problems we induce, not by the starting point. Expanding the sample
by `w_ech` gives the finite population `U`, which we treat as ground truth.

```{r build, eval = TRUE}
df_raw <- read_sav(file.path(dir_ine, "HyP_2019_Terceros.sav"))
info   <- read_sav(file.path(dir_ine, "ESTRATO_UPM_ECH2019.sav"))

df <- df_raw |>
  left_join(info |> select(numero, estrato, upm_fic), by = "numero") |>
  transmute(
    id_household  = numero,
    person_number = nper,
    strata        = estrato,
    psu           = upm_fic,
    region        = as_factor(dpto),
    sex           = ifelse(e26 == 1, "male", "female"),
    age           = e27,
    poverty       = ifelse(pobre06 == 1, 1, 0),
    w_ech         = pesoano
  ) |>
  mutate(base_weight = w_ech)

# five-year age groups, shared by sample and population
brks <- seq(0, 100, by = 5)
df   <- df |> mutate(age_grp = cut(age, breaks = brks, right = FALSE))

sample_ech_full <- df |> select(-w_ech)
U      <- df |> uncount(w_ech)
p_true <- mean(U$poverty)

# sanity check: base weights reproduce the truth before any damage
with(sample_ech_full, sum(base_weight * poverty) / sum(base_weight))  # equals p_true
```

## 3. Induce eligibility and response outcomes

Because the public file is all eligible respondents, we induce the operational
outcomes at the **household** level: a dwelling is ineligible, of unknown
eligibility, or a nonrespondent as a whole. Frame variables (region, stratum,
PSU, base weight) stay known for every household. The survey variables (sex, age,
poverty) are erased for problematic households, which collapse to a single row,
because in practice their roster is never observed.

Nonresponse is missing-at-random and, importantly, **coherent with reality**: the
poorest strata (the low-numbered ones) respond less. We assign an explicit
response level per band of strata so the gradient is concentrated where poverty
lives instead of being diluted across the whole stratum range. As poverty is
heavily concentrated in the lowest strata, this makes the poor under-represented
among respondents, exactly the direction a real survey faces.

```{r causales, eval = TRUE}
set.seed(2023)

hh <- sample_ech_full |>
  group_by(id_household) |>
  summarise(strata = first(strata), psu = first(psu), region = first(region),
            base_weight = first(base_weight), mean_age = mean(age),
            .groups = "drop")

# eligibility: ineligible ~4%, unknown ~6%, the rest eligible
u <- runif(nrow(hh))
hh$ineligible   <- as.integer(u < 0.04)
hh$unknown_elig <- as.integer(u >= 0.04 & u < 0.10)
hh$eligible     <- as.integer(u >= 0.10)

# response by stratum band: the poorest strata respond markedly less
sn <- suppressWarnings(as.numeric(as.character(hh$strata)))
strata_logit <- ifelse(sn <= 2, -0.8,
                ifelse(sn <= 5,  0.0,
                ifelse(sn <= 10, 0.9, 1.5)))
strata_logit[is.na(strata_logit)] <- 0.9
p_resp <- plogis(strata_logit + 0.040 * (hh$mean_age - 40))

hh$responded <- 0L
elig <- hh$eligible == 1
hh$responded[elig] <- rbinom(sum(elig), 1, p_resp[elig])

# respondent households keep their persons; problematic ones collapse to one row
resp_hh <- hh$id_household[hh$responded == 1 & hh$eligible == 1]
persons_resp <- sample_ech_full |>
  filter(id_household %in% resp_hh) |>
  left_join(hh |> select(id_household, ineligible, unknown_elig, responded),
            by = "id_household") |>
  select(id_household, person_number, strata, psu, region, base_weight,
         sex, age, age_grp, poverty, ineligible, unknown_elig, responded)
prob_hh <- hh |>
  filter(!(responded == 1 & eligible == 1)) |>
  transmute(id_household, person_number = 1L, strata, psu, region, base_weight,
            sex = NA_character_, age = NA_real_,
            age_grp = factor(NA, levels = levels(df$age_grp)),
            poverty = NA_real_, ineligible, unknown_elig, responded)

sample_ech <- bind_rows(persons_resp, prob_hh) |>
  arrange(id_household, person_number)
```

## 4. Build calibration targets from the population

Raking or linear calibration needs population totals for the margins. A
statistical office reads these from population projections; here we read them
from `U`. We calibrate to the number of people by five-year age group, by sex,
and by region, and we build the totals with the **same** model matrix the recipe
will use, so the column names line up.

A subtlety on totals: after dropping the ineligibles the sample represents the
**eligible** population, which is smaller than all of `U`. Since eligibility was
induced at random (independent of age, sex and region), the eligible population
has the same composition as `U` scaled by the eligible share, so we scale the
totals by that factor.

```{r targets, eval = TRUE}
elig_factor <- mean(hh$eligible)            # share of eligible households
cal_formula <- ~ age_grp + sex + region

XU     <- model.matrix(cal_formula, data = U)
totals <- colSums(XU) * elig_factor         # scaled to the eligible population
```

## 5. The weighting recipe

We pipe the whole cascade. Note the calibration step: we use **linear** distance
with `equal_within_cluster = TRUE` and `cluster = "id_household"`, which produces
a single weight per household (integrated household weighting, Lemaitre-Dufour).
This is what an official household survey wants, since every member of a dwelling
shares the same selection probability.

```{r recipe, eval = TRUE}
fitted <- weighting_spec(sample_ech, base_weights = base_weight) |>
  step_unknown_eligibility(unknown = unknown_elig == 1, by = "region",
                           cluster = "id_household") |>
  step_drop_ineligible(ineligible = ineligible == 1) |>
  step_nonresponse(respondent = responded, method = "weighting_class",
                   by = "strata", cluster = "id_household") |>
  step_calibrate(method = "linear", formula = cal_formula, totals = totals,
                 cluster = "id_household", equal_within_cluster = TRUE) |>
  step_trim_weights(method = "potter") |>
  step_round(digits = 0, method = "preserve_total") |>
  prep()

summary(fitted)
```

## 6. The estimate against the truth

We compute $\hat p$ at each stage of the cascade and compare it with the true
rate $p$. The naive (base-weight) estimate is biased downward, because the poor
responded less. The nonresponse adjustment by stratum recovers the between-strata
part of that bias, and calibration to age, sex and region recovers more.

```{r validation, eval = TRUE}
y <- sample_ech$poverty
stage_phat <- sapply(fitted$history, function(w) {
  ok <- w > 0 & !is.na(y); sum(w[ok] * y[ok]) / sum(w[ok])
})
labs <- gsub("^stage_[0-9]+_step_", "", names(stage_phat))
est  <- data.frame(stage = factor(labs, levels = labs), phat = stage_phat)

data.frame(stage = labs, phat = round(stage_phat, 5),
           dist = round(stage_phat - p_true, 5))

ggplot(est, aes(stage, phat, group = 1)) +
  geom_hline(yintercept = p_true, color = wf_amber, linewidth = 1) +
  annotate("text", x = levels(est$stage)[1], y = p_true, label = "true p",
           vjust = -0.6, hjust = 0, color = wf_amber) +
  geom_line(color = wf_lav, linewidth = 1) +
  geom_point(color = wf_violet, size = 3) +
  labs(x = NULL, y = expression(hat(p)),
       title = "Poverty rate across the weighting cascade") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Each adjustment moves the estimate toward the truth. A small residual gap
remains, which is honest and expected: the adjustments use only the frame and the
calibration margins, so the part of the bias they cannot explain is not fully
removed. Cutting the initial bias by most of its size is the realistic outcome of
weighting, not a perfect bullseye.

## 7. Per-step weight diagnostics

We inspect what each step does to the weights. First, a scatter of the weight
before versus after each step: points on the amber diagonal were left untouched,
points off it were reweighted. The integrated household calibration shows up as
aligned points, because every member of a household receives the same factor.

```{r diag-scatter, fig.height = 6}
h    <- fitted$history
labs <- gsub("^stage_[0-9]+_step_", "", names(h))

pairs <- bind_rows(lapply(seq_len(length(h) - 1L), function(i) {
  prev <- h[[i]]; cur <- h[[i + 1L]]; keep <- prev > 0 & cur > 0
  data.frame(step = factor(labs[i + 1L], levels = labs[-1]),
             prev = prev[keep], cur = cur[keep])
}))

ggplot(pairs, aes(prev, cur)) +
  geom_abline(slope = 1, intercept = 0, color = wf_amber, linewidth = 0.7) +
  geom_point(color = wf_violet, alpha = 0.2, size = 0.6) +
  facet_wrap(~ step, scales = "free", ncol = 3) +
  labs(x = "weight before the step", y = "weight after the step",
       title = "Weights before vs after each step") +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(color = wf_violet, face = "bold"))
```

Second, the distribution of the adjustment factor (the weight after the step
divided by the weight before). A factor of one (the amber line) means the step
left a unit untouched; mass away from one shows where, and how hard, the step
worked.

```{r diag-factor, fig.height = 6}
ggplot(pairs |> mutate(factor = cur / prev), aes(factor)) +
  geom_histogram(bins = 30, fill = wf_lav, color = "white") +
  geom_vline(xintercept = 1, color = wf_amber, linewidth = 0.7) +
  facet_wrap(~ step, scales = "free", ncol = 3) +
  labs(x = "adjustment factor (after / before)", y = "units",
       title = "Distribution of the adjustment factor at each step") +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(color = wf_violet, face = "bold"))
```

The Kish design effect summarises the variance cost of unequal weights. It is
useful to track it across stages as a single number per step.

```{r deff, eval = TRUE}
data.frame(
  stage = labs,
  deff  = round(vapply(h, function(w) design_effect(w)$deff, numeric(1)), 3)
)
```

## 8. Design-based inference: bootstrap and 95% CI

The variance must reflect the design (PSUs within strata) and the fact that the
recipe is re-estimated on each replicate. `bootstrap_weights()` resamples PSUs
within strata (Rao-Wu rescaling) and re-applies the whole recipe;
`boot_mean()` returns the estimate, its standard error and a 95% interval (a
normal interval around the point estimate).

```{r bootstrap, eval = TRUE}
boot <- bootstrap_weights(fitted, replicates = 200, strata = "strata", psu = "psu")
ci <- boot_mean(boot, "poverty")
ci
```

The empirical distribution of $\hat p$ across the replicates shows the sampling
variability behind that interval.

```{r boot-hist, eval = TRUE}
reps <- apply(boot$replicates, 2, function(w) {
  ok <- w > 0 & !is.na(y); sum(w[ok] * y[ok]) / sum(w[ok])
})

ggplot(data.frame(phat = reps), aes(phat)) +
  geom_histogram(bins = 30, fill = wf_lav, color = "white") +
  geom_vline(xintercept = ci$estimate, color = wf_violet, linewidth = 1) +
  geom_vline(xintercept = c(ci$ci_lower, ci$ci_upper), color = wf_violet,
             linetype = "dashed") +
  geom_vline(xintercept = p_true, color = wf_amber, linewidth = 1) +
  annotate("text", x = p_true, y = Inf, label = "true p", vjust = 1.4,
           hjust = -0.1, color = wf_amber) +
  labs(x = expression(hat(p)), y = "replicates",
       title = "Bootstrap distribution of the poverty-rate estimator",
       subtitle = "solid = estimate, dashed = 95% CI, amber = true p") +
  theme_minimal(base_size = 12)
```

## Notes

The pipeline runs on tens of thousands of records in a few seconds, so it doubles
as a light stress test. Everything is reproducible from the public ECH file plus
the seeded code, so the induced outcomes and the resulting weights are identical
on every run. The residual gap between the final estimate and the truth is a
faithful reminder that weighting reduces nonresponse bias substantially but, when
the response mechanism depends on more than the frame and the calibration
margins, does not erase it entirely.
